---
title: "Soccer Predictions Using the Bradley Terry Model"
author: "Zachary P. Malosh, MS, MBA"
date: "July 24, 2019"
output:
  pdf_document:
    toc: no
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents
\newpage

```{r setup, include=FALSE}
update_wd <- function(){
  if(toupper(basename(getwd())) == 'NOTEBOOKS'){
    setwd('..')
  }
}


knitr::opts_chunk$set(echo = FALSE)

update_wd()
source('requirements.R')
source('src/data/get_fixtures.R')
source('src/data/get_leagues.R')

library(BradleyTerry2)
```

# Included Libraries
The following libraries are used for this analysis:

+ *tidyverse*
+ *BradleyTerry2*

# League Selection
```{r 'League Selection'}
update_wd()
leagues <- get_leagues()
esp.lg <- leagues %>%
  filter(Country == 'Spain'
         & LeagueName == 'Primera Division'
         & Season == 2018)
```
The initial model here will be created using game results from La Liga in 2018. Other leagues will be pulled in to test easily changing between leagues and/or seasons. That season goes from `r esp.lg[1,]$SeasonStartDate` to `r esp.lg[1,]$SeasonEndDate`

# Getting the Games
```{r 'Get Games'}
update_wd()
esp.games.raw <- get_fixtures_by_league(esp.lg$LeagueId)
teams <- esp.games.raw %>%
  mutate(ID = HomeTeamId,
         Name = HomeTeamName,
         Logo = HomeTeamLogo,
         Venue = str_trim(str_replace(Venue, '\\(.*\\)$', '')),
         Strength = 1) %>%
  select(ID, Name, Logo, Venue, Strength) %>%
  unique() %>%
  arrange(Name)
games <- esp.games.raw %>%
  mutate(Venue = str_trim(str_replace(Venue, '\\(.*\\)$', ''))) %>%
  select(FixtureId, GameDate, Venue, HomeTeamId, HomeTeamName,
         AwayTeamId, AwayTeamName, HomeScore, AwayScore) %>%
  mutate(GameResult = ifelse(HomeScore == AwayScore, 0.5, ifelse(HomeScore > AwayScore, 1, 0)),
         GameTotal = HomeScore + AwayScore)
```
There are `r nrow(games)` games in the selected league.

## Team Data
The data found in the table below is available for each team. Strength is defaulted to 1 for all teams to begin. (Logos are omitted due to length.)

```{r 'Display Team Data'}
knitr::kable(teams %>% select(-Logo), caption = '2018 La Liga Team Data')
```

## Game Data 
For each game, the following data is available in the data set:

+ Home/Away Team ID
+ Home/Away Team Name
+ Home/Away Score
+ Game Date
+ Venue

A game result value (*GameResult*) is derived as follows:

|  Value |   Result |
|:------:|:--------:|
|      0 | Away Win |
|    0.5 |     Draw |
|      1 | Home Win |

The game total (*GameTotal*) is calculated as the sum of the home and away scores (*HomeScore + AwayScore*)

## Home Field Advantage Verification
There is no indicator for whether or not a game is at a neutral site. La Liga has a double round-robin format. Each team should have 38 games, hosting each other team once and playing each other team away once. The following table shows the number of games in a venue as the host team for each team.

```{r 'Verify Home Venues'}
homeVenues <- games %>%
  group_by(HomeTeamId, HomeTeamName, Venue) %>%
  summarise(Games = n()) %>%
  arrange(HomeTeamName, Games)
knitr::kable(homeVenues, caption = 'Team Games Hosted per Venue')
```

Based on there being 19 games hosted for each team at the same site, there are no neutral site games.

# Team Strength with Bradley-Terry Model

The Bradley-Terry pairwise comparison model was used to determine the relative team strengths. The result of the model can be found below.

```{r}
btm <- BradleyTerry2::BTm(outcome = cbind(games$HomeScore, games$AwayScore),
                          player1 = data.frame(team = games$HomeTeamName, at.home = 1),
                          player2 = data.frame(team = games$AwayTeamName, at.home = 0),
                          formula = ~team + at.home,
                          id = 'team',
                          data = games)
summary(btm)
```


