---
title: "Soccer Predictions Using the Bradley Terry Model"
author: "Zachary P. Malosh, MS, MBA"
date: "July 24, 2019"
output:
  pdf_document:
    toc: no
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents
\newpage

```{r setup, include=FALSE}
update_wd <- function(){
  if(toupper(basename(getwd())) == 'NOTEBOOKS'){
    setwd('..')
  }
}

options(stringsAsFactors = FALSE)

knitr::opts_chunk$set(echo = FALSE)

update_wd()
source('requirements.R')
source('src/data/get_fixtures.R')
source('src/data/get_leagues.R')

library(BradleyTerry2)
```

# Included Libraries
The following libraries are used for this analysis:

+ *tidyverse*

# League Selection
```{r 'League Selection'}
update_wd()
leagues <- get_leagues()
esp.lg <- leagues %>%
  filter(Country == 'Spain'
         & LeagueName == 'Primera Division'
         & Season == 2018)
```
The initial model here will be created using game results from La Liga in 2018. Other leagues will be pulled in to test easily changing between leagues and/or seasons. That season goes from `r esp.lg[1,]$SeasonStartDate` to `r esp.lg[1,]$SeasonEndDate`

# Getting the Games
```{r 'Get Games'}
update_wd()
esp.games.raw <- get_fixtures_by_league(esp.lg$LeagueId)
teams <- esp.games.raw %>%
  mutate(ID = HomeTeamId,
         Name = HomeTeamName,
         Logo = HomeTeamLogo,
         Venue = str_trim(str_replace(Venue, '\\(.*\\)$', '')),
         Strength = 1) %>%
  select(ID, Name, Logo, Venue, Strength) %>%
  unique() %>%
  arrange(Name)
games <- esp.games.raw %>%
  mutate(Venue = str_trim(str_replace(Venue, '\\(.*\\)$', ''))) %>%
  select(FixtureId, GameDate, Venue, HomeTeamId, HomeTeamName,
         AwayTeamId, AwayTeamName, HomeScore, AwayScore) %>%
  mutate(GameResult = ifelse(HomeScore == AwayScore, 0.5, ifelse(HomeScore > AwayScore, 1, 0)),
         GameTotal = HomeScore + AwayScore)
```
There are `r nrow(games)` games in the selected league.

## Team Data
The data found in the table below is available for each team. Strength is defaulted to 1 for all teams to begin. (Logos are omitted due to length.)

```{r 'Display Team Data'}
knitr::kable(teams %>% select(-Logo), caption = '2018 La Liga Team Data')
```

## Game Data 
For each game, the following data is available in the data set:

+ Home/Away Team ID
+ Home/Away Team Name
+ Home/Away Score
+ Game Date
+ Venue

A game result value (*GameResult*) is derived as follows:

|  Value |   Result |
|:------:|:--------:|
|      0 | Away Win |
|    0.5 |     Draw |
|      1 | Home Win |

The game total (*GameTotal*) is calculated as the sum of the home and away scores (*HomeScore + AwayScore*)

## Home Field Advantage Verification
There is no indicator for whether or not a game is at a neutral site. La Liga has a double round-robin format. Each team should have 38 games, hosting each other team once and playing each other team away once. The following table shows the number of games in a venue as the host team for each team.

```{r 'Verify Home Venues'}
homeVenues <- games %>%
  group_by(HomeTeamId, HomeTeamName, Venue) %>%
  summarise(Games = n()) %>%
  arrange(HomeTeamName, Games)
knitr::kable(homeVenues, caption = 'Team Games Hosted per Venue')
```

Based on there being 19 games hosted for each team at the same site, there are no neutral site games.

# Team Strength with Bradley-Terry Model

The Bradley-Terry pairwise comparison model was used to determine the relative team strengths. The result of the model can be found below.


```{r}
logisticFunction <- function(homeFieldAdvantage, homeTeamStrength, awayTeamStrength){
  return(1 / (1 + (exp(-(homeFieldAdvantage + homeTeamStrength - awayTeamStrength)))))
}

setup_games <- function(games){
  g <- data.frame(HomeTeamName = games$HomeTeamName,
                  AwayTeamName = games$AwayTeamName,
                  HomeScore = games$HomeScore,
                  AwayScore = games$AwayScore)
  draws <- g %>% filter(AwayScore == HomeScore)
  drawHomeWinReplacement <- data.frame(HomeTeamName = draws$HomeTeamName,
                                       AwayTeamName = draws$AwayTeamName,
                                       HomeScore = draws$HomeScore + 1,
                                       AwayScore = draws$AwayScore)
  drawAwayWinReplacement <- data.frame(HomeTeamName = draws$HomeTeamName,
                                       AwayTeamName = draws$AwayTeamName,
                                       HomeScore = draws$HomeScore,
                                       AwayScore = draws$AwayScore + 1)
  g <- g %>% filter(AwayScore != HomeScore)
  g <- rbind(g, drawHomeWinReplacement, drawAwayWinReplacement)
  
  g <- g %>%
    mutate(HomeMarginOfVictory = HomeScore - AwayScore,
      GameResult = ifelse(HomeMarginOfVictory > 0, 1, 0))
  return(g)
}

f <- function(p, teams, games){
  t <- teams
  x <- p
  
  teamStrengths <- x[1:(length(x)-1)]
  names(teamStrengths) <- t
  homeFieldAdvantage <- x[length(x)]
  
  g <- games %>%
    mutate(HomeStrength = teamStrengths[HomeTeamName],
      AwayStrength = teamStrengths[AwayTeamName],
      LogisticVal = logisticFunction(homeFieldAdvantage, HomeStrength, AwayStrength),
      Result = ifelse(GameResult == 1, LogisticVal, 1 - LogisticVal)
    )
  g$HomeTeamName = as.character(g$HomeTeamName)
  g$AwayTeamName = as.character(g$AwayTeamName)
  logLikelihood <- sum(log(g$Result))
  return(-1 * logLikelihood)
}

g <- setup_games(games)
teams <- sort(unique(games$HomeTeamName))
p <- rep(1, times = length(teams) + 1)
strengthOptimization <- nlm(f, p, teams = teams, games = g)
strengths <- strengthOptimization$estimate
names(strengths) <- c(teams, 'HomeFieldAdvantage')
g$HomeStrength = strengths[g$HomeTeamName]
g$AwayStrength = strengths[g$AwayTeamName]
g$LogisticVal = logisticFunction(strengths['HomeFieldAdvantage'], g$HomeStrength, g$AwayStrength)
g$LogisticResult = ifelse(g$GameResult == 1, g$LogisticVal, 1 - g$LogisticVal)
m <- lm(formula = HomeMarginOfVictory ~ LogisticResult, data = g)
coefLogisticResult <- m$coefficients['LogisticResult']
coefIntercept <- m$coefficients['(Intercept)']
g$PredictedSpread <- coefIntercept + (coefLogisticResult * g$LogisticResult)
stdDev <- summary(m)$sigma
g$HomeWinProb <- 1 - pnorm(0, mean = g$PredictedSpread, sd = stdDev)
g$PredictedResult <- ifelse(g$HomeWinProb > 0.5, 1, 0)
g$IsResultPredicted <- ifelse(g$GameResult == g$PredictedResult, 1, 0)
g$PredictionError <- g$PredictedSpread - g$HomeMarginOfVictory
g$ProbErrorSq <- (g$GameResult - g$HomeWinProb) ** 2
g$LogError <- (g$GameResult * log(g$HomeWinProb)) + ((1-g$GameResult) * log(1 - g$HomeWinProb))

benchmarks <- data.frame(RawAccuracy = mean(g$IsResultPredicted),
                         RSQ = summary(m)$r.squared,
                         RMSE = sqrt(mean(g$PredictionError ** 2)),
                         MAE = mean(abs(g$PredictionError)),
                         BrierScore = mean(g$ProbErrorSq),
                         LogLoss = -1 * mean(g$LogError))

homeFieldAdvantage <- strengths['HomeFieldAdvantage']
testHomeStrength <- strengths['Barcelona']
testAwayStrength <- strengths['Real Madrid']
testLogisticResult <- logisticFunction(homeFieldAdvantage, testHomeStrength, testAwayStrength)
testPredHomeSpread <- as.numeric(coefIntercept + (coefLogisticResult * testLogisticResult))
testPredAwaySpread <- -1 * testPredHomeSpread
testHomeWinPct <- 1 - pnorm(0.5, mean = testPredHomeSpread, sd = stdDev)
testAwayWinPct <- 1 - pnorm(0.5, mean = testPredAwaySpread, sd = stdDev)
testDrawPct <- 1 - (testHomeWinPct + testAwayWinPct)
testHomeGoalsFavored <- 1
testAwayGoalsFavored <- -1 * testHomeGoalsFavored
testSpreadHomeWinPct <- 1 - pnorm(testHomeGoalsFavored + 0.5, mean = testPredHomeSpread, sd = stdDev)
testSpreadAwayWinPct <- 1 - pnorm(testAwayGoalsFavored + 0.5, mean = testPredAwaySpread, sd = stdDev)
testSpreadDrawPct <- 1 - (testSpreadHomeWinPct + testSpreadAwayWinPct)
```


